Flow-Based Cost Allocation in Linear Optimized Power Systems are non-unique but localizable

### 2020-08-18 paper/flow_based.tex
reintegrate sparsity into capex allocation

\documentclass[11pt,twocolumn]{article}
\usepackage{graphicx}
\usepackage[left=1.80cm, right=1.80cm, top=2.00cm, bottom=2.00cm]{geometry}
\usepackage{amsmath,bm}
\usepackage[colorlinks]{hyperref}
\usepackage[backend=biber]{biblatex}
\usepackage{eurosym}
\usepackage[dvipsnames]{xcolor}
\usepackage{subcaption}
\usepackage{booktabs} % for table (toprule, bottomrule)
\usepackage{enumitem} % for alphabetical enumeration 
\usepackage{accents}
\usepackage[capitalise]{cleveref}

@@

% allocation quantities
\newcommand{\allocategeneration}[1][s, n]{A_{#1,t}}
\newcommand{\allocatestoragedispatch}[1][r, n]{A_{#1,t}}
\newcommand{\allocatepeer}[1][m \rightarrow n]{A_{#1,t}}
\newcommand{\allocateflow}[1][n]{A_{\ell,#1,t}}
\newcommand{\allocatetransaction}[1][s \rightarrow n]{A_{#1,\ell,t}}
\newcommand{\allocatestate}[1][i, n]{A_{#1,t}}
\newcommand{\allocatecost}[1][n \rightarrow i]{\mathcal{C}_{#1, t}}
\newcommand{\allocatecapexgeneration}[1][n \rightarrow s]{\mathcal{C}^{G}_{#1,t}}
\newcommand{\allocatecapexflow}[1][n \rightarrow \ell]{\mathcal{C}^{F}_{#1,t}}
\newcommand{\allocatecapexstorage}[1][n \rightarrow r]{\mathcal{C}^{E}_{#1,t}}
\newcommand{\allocateopex}[1][n \rightarrow s]{\mathcal{O}_{#1,t}}
\newcommand{\allocateemissioncost}[1][n \rightarrow s]{\mathcal{E}_{#1,t}}

\newcommand{\emission}{e_{s}}
\newcommand{\emissionprice}{\mu_{\text{CO2}}}
\newcommand{\megawatthour}{MWh$_\text{el}$}
\newcommand{\totalcost}{\mathcal{TC}}
\newcommand{\cost}{\mathcal{C}}
\newcommand{\opex}{\mathcal{O}}
\newcommand{\opexgeneration}{\mathcal{O}^G}
\newcommand{\opexflow}{\mathcal{O}^F}
\newcommand{\opexstorage}{\mathcal{O}^E}
\newcommand{\capexgeneration}{\mathcal{C}^G}
\newcommand{\capexflow}{\mathcal{C}^F}
\newcommand{\capexstorage}{\mathcal{C}^E}
\newcommand{\emissioncost}{\mathcal{E}}
\newcommand{\remainingcost}{\mathcal{R}}

@@

In a cost-optimal setup with minimized $\totalcost$, the Locational Marginal Price (LMP) describes the price for an incremental increase of electricity demand $\demand$ at node $n$. It is given by the derivative of the total system cost $\totalcost$ with respect to the local demand $\demand$
\begin{align}
\lmp = \pdv{\,\totalcost}{\,\demand}
\label{eq:lmp}
\end{align}
% 
\begin{figure}[h]
\centering
\includegraphics[width=.8\linewidth]{price_decomposition.png}
\caption{Schematic decomposition of the Locational Market Price $\lmp$. In power system model with optimal long-term operation and planning, the total system costs $\totalcost$ split into different cost terms, \ie OPEX and CAPEX for production and transmission and possibly other expenditures. }
\label{fig:price_decomposition}
\end{figure}
% 
This leads to a nodal pricing where over the span of optimized timesteps $t$, the system costs are partially or totally payed back by the consumers 
\begin{align}
\totalcost - \remainingcost &=  \sum_{n,t} \lmp \, \demand
\label{eq:total_cost_reduced}
\end{align}
depending on the costs $\remainingcost$ which are independent of the nodal demand  
\begin{align}
 \pdv{\remainingcost}{\demand} = 0
\end{align}
% 
Generally speaking, the cost term $\remainingcost$, not covered by the consumers, results from additional demands on the network design, such as capacity expansion limits or minimum share of one technology in the power mix. However, in most cases, where $\remainingcost \ll  \totalcost$, these play a minor role.

@@

% where $\costfactor$ is either a fixed operational price or a composition of shadowprices
Following the implications of \cref{eq:lmp,eq:total_cost_reduced},  we define the cost $\allocatecost$ that consumers at bus $n$ have pay to asset $i$ at time $t$, in order to compensate for $\cost_i$, as  
\begin{align}
    \allocatecost = \costfactor \,\pdv{\state}{\demand} \demand
    \label{eq:cost_allocation}
\end{align}
The product on the right hand side 
\begin{align}
    \allocatestate = \pdv{\state}{\demand} \demand
    \label{eq:state_allocation}
\end{align}    
may be interpreted as the amount of power that asset $i$ supplies demand $\demand$ with. It heavily relies on the derivative of the operational state with respect to the nodal demand $\partial \state / \partial \demand$.
From the natural fact that the sum of all contributions must return the cost term, 
\begin{align}
    \cost_i = \sum_{n,t} \allocatecost
    \label{eq:cost_payback}
\end{align}
it follows that $\allocatestate$ must fulfill
\begin{align}
    \state = \sum_n \allocatestate
    \label{eq:state_allocation_constraint}
\end{align}
The last equation states that all power produced or processed by asset $i$ must be totally consumed by the network demand $\demand$.  
Finally, the total contribution from node $n$ at time $t$ to the cost term $\cost$ amounts 
\begin{align}
    \cost_{n,t} = \sum_i \allocatecost
\end{align}
\end{subequations}

@@

Let the operational price for an asset $i$ be given by $o_i$. Then, for example the OPEX occasioned by generator $s$ is given by 
\begin{align}
    \opexgeneration_s = \sum_t \operationalpricegeneration \, \generation 
    \label{eq:opexgeneration}
\end{align}
where $\generation$ denotes its power generation at time $t$. As \cref{eq:opexgeneration} matches the form of \cref{eq:cost_decomposition} which allows us to use the above presented scheme in \cref{eq:general_scheme}. As a result we obtain 
\begin{align}
    \allocateopex[n \rightarrow s] &= 
   \operationalpricegeneration \,  \allocategeneration
\label{eq:allocate_opexGeneration_detailed}
\end{align}
which is the contribution of $\demand$ to the OPEX at generator $s$.
The quantity  
\begin{align}
 \allocategeneration = \pdv{\generation}{\demand} \, \demand
 \label{eq:allocate_peer}
\end{align}
can be considered as the power that is produced by generator $s$ and consumed at node $n$ at time $t$. \\
In the same manner, we can follow the scheme allocate OPEX for flow $\flow$ and storage unit dispatch $\storagedispatch$. As we assume a bidirectional flow on line $\ell$, the OPEX is set proportional to the absolute value of the flow. The upper section in \cref{tab:cost_allocation_map} shows the mapping of variables to \cref{eq:general_scheme} in order to define the full OPEX allocation.

@@

\subsection{Allocate Emission Expenditures}

Given a fix price for emissions in CO$_2$ equivalents $\emissionprice$, the cost term for emission adds up to 
\begin{align}
 \emissioncost = \emissionprice \, \sum_s  \emission \, \generation
\end{align}
where $\emission$ denotes the emission factor in tonne-CO$_2$ per \megawatthour for generator $s$.
We use the same set of equation as given through \cref{eq:general_scheme,eq:opex_allocation} but update the cost factor to $\costfactor \Rightarrow \emissionprice \, \emission$. For consumers at bus $n$ at time $t$ this leads to the emission cost of 
\begin{align}
 \allocateemissioncost = \emissionprice \, \emission \, \allocategeneration
\end{align}
associated with generator $s$.

\subsection{Capacity Expansion Limit}

In more realistic setups, generators, lines or other assets can only be built up to a certain limit. This might caused by land use restrictions or social acceptance problems. %In the following give a general solution how the cost allocation may handle this.  
However, when constraining the capacity $\capacity$  for a subset $I$ of assets to an upper limit,
\begin{align}
    \capacity - \capacityupper \le 0 \resultsin{\muuppernom} \hpad \Forall{i \in I}
\label{eq:capacityexpansionmaximum}
\end{align}
the zero profit condition for asset $i$ alters. Then the revenues don't match the expenditures as soon as the constraint becomes binding. More precisely, the revenue from the shadow prices given in \cref{tab:cost_allocation_map} exceed the CAPEX $\cost_i$ by $\remainingcost_i$,  where  $\remainingcost_i$ is given by
\begin{align}
    \remainingcost_i = - \muuppernom \capacity
    \Forall{i}
\end{align}
% \begin{align}
%  \cost_i - \remainingcost_i =  \sum_t \costfactor \, \state \Forall{i}
%  \label{eq:capex_generation_duality_bf2}
% \end{align}
For the cost allocation, these means the prices in \cref{tab:cost_allocation_map} have to be adjusted, as otherwise consumers would pay too much. As we show in ... updating the prices by a fix ratio 
\begin{align}
    \costfactor \rightarrow \dfrac{c_i}{c_i + \muuppernom}\, \costfactor
    \label{eq:adjustment_lowernom}
\end{align}
will ensure that the allocated CAPEX sum up to the actual CAPEX at asset $i$, meaning that \cref{eq:cost_payback} holds. Note that $\muuppernom$ is a positive number, so the adjustment cannot increase the paid price.

\subsection{Brownfield Constraints}

In order to take already built infrastructure into account, the capacity $\capacity$ can be constrained to a minimum required capacity $\capacitylower$. Mathematically this translates to 
\begin{align}
    \capacitylower - \capacity  \le 0 \resultsin{\mulowernom} \hpad \forall{i \in I}
\label{eq:capacityexpansionminimum}
\end{align}
Again, such a setup alters the zero profit condition of asset $i$, as soon as the constraint becomes binding. 
In that case, asset $i$ does not collect enough revenue in order to match the CAPEX. The difference is given by 
\begin{align}
    \remainingcost_i = \mulowernom \capacity
    \Forall{i}
\end{align}
In constrast to 
Again, we can balance this effect out by updating the cost factor to 
\begin{align}
    \costfactor \rightarrow \costfactor +  \dfrac{\mulowernom \capacity}{\sum_t \state}
\end{align}
As $\mulowernom$ is a positive number, this will increase the price and therefore the revenue for asset $i$.

@@

We apply the cost allocation on a German power system model

>>>>

\documentclass[11pt,twocolumn]{article}
\usepackage{graphicx}
\usepackage[left=1.80cm, right=1.80cm, top=2.00cm, bottom=2.00cm]{geometry}
\usepackage{amsmath,bm}
\usepackage[colorlinks]{hyperref}
\usepackage[backend=biber, bibencoding=utf8]{biblatex}
\usepackage{eurosym}
\usepackage[dvipsnames]{xcolor}
\usepackage{subcaption}
\usepackage{booktabs} % for table (toprule, bottomrule)
\usepackage{enumitem} % for alphabetical enumeration 
\usepackage{accents}
\usepackage[capitalise]{cleveref}

@@

% Cost Terms
\newcommand{\emission}{e_{s}}
\newcommand{\emissionprice}{\mu_{\text{CO2}}}
\newcommand{\megawatthour}{MWh$_\text{el}$}
\newcommand{\totalcost}{\mathcal{TC}}
\newcommand{\cost}{\mathcal{C}}
\newcommand{\opex}{\mathcal{O}}
\newcommand{\opexgeneration}{\mathcal{O}^G}
\newcommand{\opexflow}{\mathcal{O}^F}
\newcommand{\opexstorage}{\mathcal{O}^E}
\newcommand{\capexgeneration}{\mathcal{C}^G}
\newcommand{\capexflow}{\mathcal{C}^F}
\newcommand{\capexstorage}{\mathcal{C}^E}
\newcommand{\emissioncost}{\mathcal{E}}
\newcommand{\remainingcost}{\mathcal{R}}
\newcommand{\scarcitycost}{\remainingcost^\text{scarcity}}
\newcommand{\subsidycost}{\remainingcost^\text{subsidy}}

% allocation quantities
\newcommand{\allocategeneration}[1][s, n]{A_{#1,t}}
\newcommand{\allocatestoragedispatch}[1][r, n]{A_{#1,t}}
\newcommand{\allocatepeer}[1][m \rightarrow n]{A_{#1,t}}
\newcommand{\allocateflow}[1][n]{A_{\ell,#1,t}}
\newcommand{\allocatetransaction}[1][s \rightarrow n]{A_{#1,\ell,t}}
\newcommand{\allocatestate}[1][i, n]{A_{#1,t}}

\newcommand{\allocatecost}[1][n \rightarrow i]{\cost_{#1, t}}
\newcommand{\allocatecapexgeneration}[1][n \rightarrow s]{\capexgeneration_{#1,t}}
\newcommand{\allocatecapexflow}[1][n \rightarrow \ell]{\capexflow_{#1,t}}
\newcommand{\allocatecapexstorage}[1][n \rightarrow r]{\capexstorage_{#1,t}}
\newcommand{\allocateopex}[1][n \rightarrow s]{\opex_{#1,t}}
\newcommand{\allocateemissioncost}[1][n \rightarrow s]{\emissioncost_{#1,t}}
\newcommand{\allocatescarcitycost}[1][n \rightarrow i]{\scarcitycost_{#1,t}}
\newcommand{\allocatesubsidycost}[1][n \rightarrow i]{\subsidycost_{#1,t}}

@@

In a cost-optimal setup with minimized $\totalcost$, the Locational Marginal Price (LMP) describes the price for an incremental increase of electricity demand $\demand$ at node $n$. It is given by the derivative of the total system cost $\totalcost$ with respect to the local demand $\demand$
\begin{align}
\lmp = \pdv{\,\totalcost}{\,\demand}
\label{eq:lmp}
\end{align}
% 
\begin{figure}[h]
\centering
\includegraphics[width=.8\linewidth]{price_decomposition.png}
\caption{Schematic decomposition of the Locational Market Price $\lmp$. In power system model with optimal long-term operation and planning, the total system costs $\totalcost$ split into different cost terms, \ie OPEX and CAPEX for production and transmission and possibly other expenditures. }
\label{fig:price_decomposition}
\end{figure}
% 
This leads to a nodal pricing where over the span of optimized timesteps $t$, the system costs are partially or totally payed back by the consumers 
\begin{align}
\totalcost - \remainingcost &=  \sum_{n,t} \lmp \, \demand
\label{eq:total_revenue}
\end{align}
depending on the costs $\remainingcost$ which are independent of the nodal demand  
\begin{align}
 \pdv{\remainingcost}{\demand} = 0
\end{align}
% 
Generally speaking, the cost term $\remainingcost$, not covered by the consumers, results from additional demands on the network design, such as capacity expansion limits or minimum share of one technology in the power mix. However, in most cases, where $\remainingcost \ll  \totalcost$, these play a minor role.

@@

% where $\costfactor$ is either a fixed operational price or a composition of shadowprices
Following the implications of \cref{eq:lmp,eq:total_revenue},  we define the cost $\allocatecost$ that consumers at bus $n$ have pay to asset $i$ at time $t$, in order to compensate for $\cost_i$, as  
\begin{align}
    \allocatecost = \costfactor \,\pdv{\state}{\demand} \demand
    \label{eq:cost_allocation}
\end{align}
The product on the right hand side 
\begin{align}
    \allocatestate = \pdv{\state}{\demand} \demand
    \label{eq:state_allocation}
\end{align}    
may be interpreted as the amount of power that asset $i$ supplies demand $\demand$ with. It heavily relies on the derivative of the operational state with respect to the nodal demand $\partial \state / \partial \demand$.
From the natural fact that the sum of all contributions must return the cost term, 
\begin{align}
    \cost_i = \sum_{n,t} \allocatecost
    \label{eq:cost_payback}
\end{align}
it follows that $\allocatestate$ must fulfill
\begin{align}
    \state = \sum_n \allocatestate
    \label{eq:state_allocation_constraint}
\end{align}
The last equation states that all power produced or processed by asset $i$ must be totally consumed by the network demand $\demand$.  
Finally, the total contribution from node $n$ at time $t$ to the cost term $\cost$ amounts 
\begin{align}
    \cost_{n,t} = \sum_i \allocatecost
\end{align}
\end{subequations}

@@

Let the operational price for an asset $i$ be given by $o_i$. Then, for example the OPEX occasioned by generator $s$ is given by 
\begin{align}
    \opexgeneration_s = \sum_t \operationalpricegeneration \, \generation 
    \label{eq:opexgeneration}
\end{align}
where $\generation$ denotes its power generation at time $t$. As \cref{eq:opexgeneration} matches the form of \cref{eq:cost_decomposition} which allows us to use the above presented scheme in \cref{eq:general_scheme}. As a result we obtain 
\begin{align}
    \allocateopex[n \rightarrow s] &= 
   \operationalpricegeneration \,  \allocategeneration
\label{eq:allocate_opexGeneration_detailed}
\end{align}
which is the contribution of $\demand$ to the OPEX at generator $s$.
The quantity  
\begin{align}
 \allocategeneration = \pdv{\generation}{\demand} \, \demand
 \label{eq:allocate_peer}
\end{align}
can be considered as the power that is produced by generator $s$ and consumed at node $n$ at time $t$. \\
In the same manner, we can follow the scheme allocate OPEX for flow $\flow$ and storage unit dispatch $\storagedispatch$. As we assume a bidirectional flow on line $\ell$, the OPEX is set proportional to the absolute value of the flow. The upper section in \cref{tab:cost_allocation_map} shows the mapping of variables to \cref{eq:general_scheme} in order to define the full OPEX allocation. \\

The scheme works for all other cost attached to the operational state of an asset $i$. Given for example a fix price for emissions $\emissionprice$ in \euro\, per tonne-CO$_2$ equivalents, the cost term for emission adds up to 
\begin{align}
 \emissioncost = \emissionprice \, \sum_s  \emission \, \generation
\end{align}
where $\emission$ denotes the emission factor in tonne-CO$_2$ per \megawatthour\, of generator $s$.
The allocated payment for consumers at bus $n$ at time $t$ assigned to generator $s$ is then given by 
\begin{align}
 \allocateemissioncost = \emissionprice \, \emission \, \allocategeneration
\end{align}

@@

\subsection{Design Constraints}

Power system modelling does rarely follow a pure Greenfield approach with unlimited capacity expansion. Rather, today's models are setting various constraints defining socio-political or  technical requirements. As mentioned before this will alter the equality of total cost and total revenue, \ie leads to $\remainingcost \ne 0$ in \cref{eq:total_revenue}. More precisely, each constraint $h_j$ (other then the nodal balance constraint) of the form 
\begin{align}
    h_j \left(\state, \capacity \right) - K < 0
\end{align}
where $K$ is any non-zero constant, will result in a cost term contributing to $\remainingcost$ and in some cases alter \cref{eq:cost_decomposition} to 
\begin{align}
    \cost_i - \remainingcost_i = \sum_t \costfactor \, \state
\end{align}
In the following we highlight two often used classes of constraints and show how to incorporate them into the cost allocation.

\subsubsection{Capacity Expansion Limit}

In more realistic setups, generators, lines or other assets can only be built up to a certain limit. This might be due to land use restrictions or social acceptance problems. %In the following give a general solution how the cost allocation may handle this.  
However, when constraining the capacity $\capacity$  for a subset $I$ of assets to an upper limit $\capacityupper$, in the form of 
\begin{align}
    \capacity - \capacityupper \le 0 \resultsin{\muuppernom} \Forall{i \in I}
\label{eq:capacityexpansionmaximum},
\end{align}
the zero profit condition alters as soon as the constraint becomes binding. Then, the revenue of asset $i$ exceeds its total expenditures (OPEX + CAPEX). More precisely, the allocated CAPEX in \cref{tab:cost_allocation_map} will surpass the actual CAPEX of asset $i$ by the cost it has to pay for the scarcity, given by the absolute value of 
\begin{align}
    \scarcitycost_i = - \muuppernom \capacity \Forall{i \in I}
\end{align}
% \begin{align}
%  \cost_i - \remainingcost_i =  \sum_t \costfactor \, \state \Forall{i}
%  \label{eq:capex_generation_duality_bf2}
% \end{align}

% In order to ensure that the allocated CAPEX sum up to the actual CAPEX at asset $i$, meaning that \cref{eq:cost_payback} holds, we adjust the CAPEX allocations to 
% \begin{align}
%     \allocatecost = \left(\dfrac{c_i}{c_i + \muuppernom}\right) \, \costfactor \, \allocatestate \Forall{i \in I}
% \end{align}
% with the cost factors $\costfactor$ for CAPEX given in \cref{tab:cost_allocation_map}. 
% The costs which consumers at $n$ have to pay for the scarcity impacting asset $i$ are given   
% \begin{align}
%     \allocatescarcitycost = \left(\dfrac{\muuppernom}{c_i + \muuppernom}\right) \, \costfactor \, \allocatestate \Forall{i \in I}
% \end{align}

The costs which consumers at $n$ have to pay for the scarcity impacting asset $i$ are given by 
\begin{align}
    \allocatescarcitycost = \dfrac{\muuppernom}{c_i + \muuppernom} \, \allocatecost^I \Forall{i \in I}
\end{align}
where $\allocatecost^I$ denotes the CAPEX allocation presented above.

% have to be adjusted, as otherwise consumers would pay too much. As we show in ... updating the prices by a fix ratio 
% \begin{align}
%     \costfactor \rightarrow \dfrac{c_i}{c_i + \muuppernom}\, \costfactor
%     \label{eq:adjustment_lowernom}
% \end{align}
% will . Note that $\muuppernom$ is a positive number, so the adjustment cannot increase the paid price.

\subsubsection{Brownfield Constraints}

In order to take already built infrastructure into account, the capacity $\capacity$ can be constrained to a minimum required capacity $\capacitylower$. Mathematically this translates to 
\begin{align}
    \capacitylower - \capacity  \le 0 \resultsin{\mulowernom} \hpad \forall{i \in I}
\label{eq:capacityexpansionminimum}
\end{align}
Again, such a setup alters the zero profit condition of asset $i$, as soon as the constraint becomes binding. 
In that case, asset $i$ does not collect enough revenue in order to match the CAPEX. The difference, given by 
\begin{align}
    \subsidycost_i = \mulowernom \capacity
    \Forall{i}
\end{align}
has to be subsidized by governments or communities. It is rather futile wanting to allocate these cost to consumers as assets may not gain any revenue for their operational state, \ie where $\cost^I = \subsidycost_i $.

% Again, we can balance this effect out by updating the cost factor to 
% \begin{align}
%     \costfactor \rightarrow \costfactor +  \dfrac{\mulowernom \capacity}{\sum_t \state}
% \end{align}
% In constrast to \cref{eq:adjustment_lowernom} the price is not weighted by a factor but shifted for all time steps. This is required as 
% As $\mulowernom$ is a positive number, this will increase the price and therefore the revenue for asset $i$.

@@

In order to showcase the cost allocation at a more complex system, we apply it to an cost-optimized German power system model with 50 nodes and one year time span with hourly resolution. The model builds on the PyPSA-EUR workflow \cite{horsch_pypsa-eur_2020} with technigal details and assumptions reported in \cite{horsch_pypsa-eur_2018}. We follow a browfield approach where transmission lines start from today's capacity values originally retreived from the ENTSO-E Transmission System Map \cite{entsoe-e_entso-e_nodate}. Pre-installed generation capacities for wind and solar are distributed proportionally to the average power potential at each site excluding those with an average capacity factor of 10\%. Further, wind and solar capacity expansion are limited by land use restriction. These consider agriculture, urban, forested and protected areas based on the CORINE and NATURA2000 database \cite{corine2012,natura2000}. Pumped Hydro Storages (PHS) are fixed to today's capacities with no more expansion allowed. Additionally, unlimited batteries and H$_{2}$-storages are allowed at each node. We impose a effective carbon price of 120 \euro per tonne-CO$_{2}$.

\Cref{fig:network} shows the estimate of existent capacities on the left hand side and the capacity expansions on the right. 
\begin{figure*}
    \centering
    \includegraphics[width=\linewidth]{de50bf}
    \label{fig:network}
\end{figure*}
